---
title: "Network Science Final project"
author: "Ravi Brenner"
date: "2025-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r}
library(igraph)
library(tidyverse)
library(patchwork)
library(latex2exp)
theme_set(theme_minimal())
```


Read in the data, filtering to trees with size >= 20 and at least 2 generations of spread
```{r}
trees <- readRDS("data_tibble_trees.RDS") |>
  janitor::clean_names() |>
  filter(size >= 20,
         generations >= 2) |>
  mutate(tree = map(tree,\(x) igraph::upgrade_graph(x)),
         full_title = str_c(disease,", ",str_replace_na(sublocation,""), " ",country,", ", year))
```

To add in here: random and targeted breakdown of each network.

Maybe? Calculate dispersion like they do in the paper? might be more complicated than it's worth
```{r}
trees$avg_out_degree <- NA
trees$avg_non_terminal_out_degree <- NA
trees$gen1 <- NA
trees$avg_gen01_out_degree <- NA
trees$ss_threshold <- NA
trees$ss_count <- NA
trees$ss_threshold_non_term <- NA
trees$ss_non_term_count <- NA
trees$ss_threshold_gen01 <- NA
trees$ss_gen01_count <- NA

for (i in 1:nrow(trees)){
  current_tree <- trees$tree[[i]]
  
  # average out degree of all nodes
  trees$avg_out_degree[i] <- mean(degree(current_tree,mode = "out"))
  
  # average out degree of non-terminal nodes
  non_terminal_nodes <- V(current_tree)[degree(current_tree, mode = "out") > 0]
  trees$avg_non_terminal_out_degree[i] <- mean(degree(current_tree, v=non_terminal_nodes,mode = "out"))
  
  # first gen size
  first_node <- V(current_tree)[degree(current_tree, mode = "in") == 0]
  first_gen_nodes <- V(current_tree)[distances(current_tree,v = first_node) == 1]
  trees$gen1[i] <- length(first_gen_nodes)
  trees$avg_gen01_out_degree[i] <- mean(degree(current_tree, v = c(first_node,first_gen_nodes), mode = "out"))
  
  # superspreaders 
  # "Lloyd-Smith and colleagues [3] quantitatively defined superspreaders as cases that cause more secondary infections than the 99th percentile of a Poisson(R0) distribution, where R0 is the basic reproductive number, or average number of secondary infections per case."
  trees$ss_threshold[i] <- qpois(0.99,lambda = trees$avg_out_degree[i])
  ss_nodes <- V(current_tree)[degree(current_tree, mode = "out") > trees$ss_threshold[i]]
  trees$ss_count[i] <- length(ss_nodes)
  
  # superspreaders, calculating threshold excluding terminal nodes
  trees$ss_threshold_non_term[i] <- qpois(0.99, lambda = trees$avg_non_terminal_out_degree[i])
  ss_nodes <- V(current_tree)[degree(current_tree, mode = "out") > trees$ss_threshold_non_term[i]]
  trees$ss_non_term_count[i] <- length(ss_nodes)
  
  # superspreaders, calculating threshold using gen 0 and 1 only
  trees$ss_threshold_gen01[i] <- qpois(0.99, lambda = trees$avg_gen01_out_degree[i])
  ss_nodes <- V(current_tree)[degree(current_tree, mode = "out") > trees$ss_threshold_gen01[i]]
  trees$ss_gen01_count[i] <- length(ss_nodes)
  
  
  # Try to visualize the trees?
  # plot(trees$tree[[i]],layout = layout_with_kk,
  #      main = trees$full_title[[i]],
  #      vertex.size = sqrt(igraph::degree(trees$tree[[i]])),
  #      edge.arrow.size=0.4,
  #      vertex.label.cex = 0.5)
}
```

Number of trees of each disease 

Note: they limit most future analyses to diseases with at least 3 trees. We could do the same?

Those diseases are: Ebola, COVID-19, MERS, norovirus, SARS, and Measles
```{r}
p.tree_count <- trees |>
  count(disease) |>
  ggplot(aes(y = reorder(disease,n), x = n)) + 
  geom_col(fill = "lightblue") + 
  labs(x = "Number of trees",
       y = "Disease") + 
  theme(text = element_text(size = 15))
p.tree_count
```

Size of trees
```{r}
p.tree_size <- trees |>
  mutate(.by = disease, n = n()) |>
  ggplot(aes(x = size, y = reorder(disease,n))) +
  geom_point() + 
  labs(x = "Tree size",
       y = "Disease") + 
  theme(text = element_text(size = 15))
p.tree_size
```

```{r}
p.tree_count + p.tree_size + 
  plot_layout(axes = "collect") 

ggsave("number_size_plot.png",width = 10, height = 6, bg="white")
```

Look at the biggest trees, so that we can visualize a few of them
```{r}
trees |> 
  arrange(-size) |>
  select(id,size) |>
  head(10)
```

Plotting an individual tree, can use this in the slides to explain the terminal vs. 1st gen concept, how that impacts the average out degree / R0 calculation
```{r}
# only need to change this to get a new set of plots
tree_id <- "usa.2020.covid.6.00"

plot_tree <- trees[which(trees$id==tree_id), ]$tree[[1]]  

tree_title <- trees |>
  filter(id == tree_id) |>
  select(full_title) |>
  pull()

# Mark all nodes in black
V(plot_tree)$color <- "black"

# Mark first node in red
V(plot_tree)[degree(plot_tree, mode = "in") == 0]$node_type <- "first"
V(plot_tree)[degree(plot_tree, mode = "in") == 0]$color <- "red"

# Mark gen 1 nodes in yellow
first_node <- V(plot_tree)[degree(plot_tree, mode = "in") == 0]
V(plot_tree)[distances(plot_tree, v=first_node)==1]$node_type <- "gen1"
V(plot_tree)[distances(plot_tree, v=first_node)==1]$color <- "yellow"

# Mark terminal nodes in blue
V(plot_tree)[degree(plot_tree, mode = "out") == 0]$node_type <- "terminal"
V(plot_tree)[degree(plot_tree, mode = "out") == 0]$color <- "blue"

# Mark nodes that are terminal and first gen in green
V(plot_tree)[degree(plot_tree, mode = "out") == 0 & distances(plot_tree, v=first_node)==1]$node_type <- "terminal and gen1"
V(plot_tree)[degree(plot_tree, mode = "out") == 0 & distances(plot_tree, v=first_node)==1]$color <- "lightgreen"
par(mfrow=c(1,1))

plot(plot_tree, layout = layout_with_kk,
     vertex.color = V(plot_tree)$color, 
     vertex.size=5, vertex.label="", 
     edge.arrow.size=.15, edge.width=.2,edge.color="black",
     main = tree_title)

legend("bottomright",
       legend=c("Index node", "1st gen","Terminal node","Terminal 1st gen", "All other nodes"),  
       fill = c("red","yellow","blue","lightgreen","black") ,
       cex = 0.7)

# another version in a circle
plot(plot_tree, layout = layout_as_tree(plot_tree,circular = TRUE),
     vertex.color = V(plot_tree)$color, vertex.size=5, vertex.label="", 
     edge.arrow.size=.15, edge.width=.2,edge.color="black",
     main = tree_title)

legend("topright",
       legend=c("Index node", "1st gen","Terminal node","Terminal 1st gen", "All other nodes"),  
       fill = c("red","yellow","blue","lightgreen","black") ,
       cex = 0.7)

# Can maybe look at which nodes have high degree this way?
degree_table <- tibble(nodes = as.numeric(V(plot_tree)),
                       out_degree = degree(plot_tree, mode = "out"),
                       node_type = V(plot_tree)$node_type)

ggplot(degree_table) +
  geom_point(aes(x = nodes, y = out_degree, color = node_type))
```


plot degree/R0 by disease, and show how it changes depending on which nodes you include.

```{r}
trees |> 
  pivot_longer(cols = c(avg_out_degree,avg_non_terminal_out_degree, avg_gen01_out_degree),
               names_to = "degree_type",
               values_to = "avg_degree") |>
  ggplot(aes(y = disease, x = avg_degree, color = degree_type)) + 
  geom_boxplot()  +
  scale_color_discrete(name = "Nodes included",
                     labels = c("Gen 0 and 1 nodes","Non-terminal nodes","All nodes")) + 
  scale_x_continuous(n.breaks = 12,
                     minor_breaks = NULL) +
  labs(x = TeX("$R_{0}$ / Average out degree"),
       y = "") +
  theme(text = element_text(size = 15),
        legend.position = "bottom",
        legend.text = element_text(size = 12, hjust = -10))

ggsave("r0_by_disease.png", width = 8, heigh = 6, bg = "white")
```

Plot superspreader cutoffs by disease
```{r}
p.ss_threshold <- trees |> 
  pivot_longer(cols = c(ss_threshold,ss_threshold_non_term, ss_threshold_gen01),
               names_to = "ss_threshold",
               values_to = "avg_degree") |>
  ggplot(aes(y = disease, x = avg_degree, color = ss_threshold)) + 
  geom_boxplot()  +
  scale_color_manual(name = "Nodes included",
                     labels = c("All nodes","Gen 0 and 1 nodes","Non-terminal nodes"),
                     values = c("#619CFF","#F8766D","#00BA38")) +
  labs(x = "Superspreader threshold",
       y = "Disease") +
  theme(legend.position = "bottom")

p.ss_threshold
```

Plot Percent of superspreaders  by disease
```{r}
p.pct_ss <- trees |> 
  mutate(ss_count = ss_count / size,
         ss_non_term_count = ss_non_term_count / size,
         ss_gen01_count = ss_gen01_count / size) |>
  pivot_longer(cols = c(ss_count,ss_non_term_count, ss_gen01_count),
               names_to = "node_type",
               values_to = "ss_count") |>
  ggplot(aes(y = disease, x = ss_count, color = node_type)) + 
  geom_boxplot()  +
  scale_color_manual(name = "Nodes included",
                     labels = c("All nodes","Gen 0 and 1 nodes","Non-terminal nodes"),
                     values = c("#619CFF","#F8766D","#00BA38")) +
  labs(x = "Percent of superspreaders",
       y = "Disease") +
  theme(legend.position = "bottom")
p.pct_ss

p.ss_threshold + p.pct_ss + 
  plot_layout(axes = "collect", guides = "collect") &
  theme(legend.position = "bottom",
        text = element_text(size = 15))

ggsave("pct_ss_by_disease.png",width = 12, height = 6, bg="white")
```

Plot R0 vs percent of superspreaders

Sort of interesting that the percent of superspreaders is highest for lower R0 values. Makes sense since the threshold increases with increasing R0 values, and a higher threshold becomes harder to achieve. Heuristically, if you have a high R0 disease like measles (r0 ~= 12), then you need to spread to a LOT of people to be considered a super spreader.
```{r}
p.r0_ss_nonterm <- trees |>
   mutate(ss_count = ss_count / size,
         ss_non_term_count = ss_non_term_count / size,
         ss_gen01_count = ss_gen01_count / size) |>
  ggplot(aes(x = avg_non_terminal_out_degree, y = ss_non_term_count,
             color = disease, label = disease)) + 
  geom_point() + 
  geom_text(size = 3, check_overlap = TRUE, hjust = -0.1, vjust = -0.5) + 
  scale_x_continuous(limits = c(1,13),
                     n.breaks = 13) +
  theme(legend.position= "none") +
  labs(x = "R0, non-terminal nodes",
       y = "Percent of superspreaders") 

p.r0_ss_nonterm
```


```{r}
p.r0_ss_gen01 <- trees |>
   mutate(ss_count = ss_count / size,
         ss_non_term_count = ss_non_term_count / size,
         ss_gen01_count = ss_gen01_count / size) |>
  ggplot(aes(x = avg_gen01_out_degree, y = ss_gen01_count,
             color = disease, label = disease)) + 
  geom_point() + 
  geom_text(size = 3, check_overlap = TRUE, hjust = -0.1, vjust = -0.5) + 
  scale_x_continuous(limits = c(1,10),
                     n.breaks = 10) + 
  theme(legend.position= "none") +
  labs(x = "R0, Gen 0 and 1 nodes",
       y = "Percent of superspreaders") 
p.r0_ss_gen01
```


```{r}
p.r0_ss_nonterm + p.r0_ss_gen01 + 
  plot_layout(axes = "collect", guides = "collect") &
  theme(legend.position = "none",
        text = element_text(size = 15))

ggsave("r0_ss_scatter.png",width = 12, height = 6, bg="white")
```




